A Fully Abstract Game Semantics for Countable Nondeterminism

### 2018-11-08 cones/paper/lipics-v2018-authors/sources/cones.tex
Haven't made a commit for ages.

\acknowledgements{This material is based on work supported by the EPSRC under Grant No.~EP/K018868/1.  
I am grateful to Martin Hyland for our conversation that helped me to develop some of this material.}%optional

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{Computer Science Logic 2018 (CSL 2018)}
\EventShortTitle{CSL 2018}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{September 4 -- 7, 2018}
\EventLocation{Oxford, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%\nolinenumbers %uncomment to disable line numbering
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\nat}{{\mathtt{nat}}}
\newcommand*\from{\colon}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\pred}{{\mathtt{pred\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\com}{{\mathtt{com}}}
\newcommand{\skipp}{{\mathsf{skip}}}
\newcommand{\neww}{{\mathsf{new}}}
\newcommand{\mkvar}{{\mathsf{mkvar}\;}}
\newcommand{\deref}{\texttt{@}}
\newcommand{\ia}[2]{\langle #1 , #2 \rangle}
\newcommand{\stup}[3]{\langle #1 \mid #2 \mapsto #3 \rangle}
\newcommand{\Var}{{\mathtt{Var}}}
\newcommand{\new}{{\mathtt{new}}}
\newcommand{\case}{{\mathtt{case}}}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\bN}{\mathbb{N}}
\newcommand{\bC}{\mathbb{C}}
\newcommand{\deno}[1]{\left\llbracket#1\right\rrbracket}
\newcommand{\converges}{\Downarrow}
\newcommand{\mustconverge}{\converges^{\text{must}}}
\newcommand{\Iflt}{\mathtt{If{<}\;}}
\newcommand{\Ifgt}{\mathtt{If{>}\;}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{\lmam}{\sqsubseteq_{m\&m}}
\newcommand{\emam}{\equiv_{m\&m}}
\newcommand{\lst}{\lesssim}
\newcommand{\rulename}[1]{\LeftTirNameStyle{#1}}
\newcommand{\ts}{{\;\vdash}}
\newcommand{\nts}{{\;\not\vdash}}
\newcommand{\OP}{\{O,P\}}
\newcommand{\QA}{\{Q,A\}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\esuchthat}{\;.\;}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\tensor}{\otimes}
\renewcommand{\implies}{\multimap}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\s}{\mathfrak s}
\newcommand{\C}{{\mathcal{C}}}
\newcommand{\G}{\mathcal G}
\DeclareMathOperator{\Int}{int}
\newcommand{\dv}{{\lightning}}
\DeclareMathOperator{\pocl}{pocl}
\newbox\gnBoxA
\newdimen\gnCornerHgt
\setbox\gnBoxA=\hbox{$\ulcorner$}
\global\gnCornerHgt=\ht\gnBoxA
\newdimen\gnArgHgt
\def\pv #1{%
    \setbox\gnBoxA=\hbox{$#1$}%
    \gnArgHgt=\ht\gnBoxA%
    \ifnum     \gnArgHgt<\gnCornerHgt \gnArgHgt=0pt%
    \else \advance \gnArgHgt by -\gnCornerHgt%
    \fi \raise\gnArgHgt\hbox{$\ulcorner$} \box\gnBoxA %
    \raise\gnArgHgt\hbox{$\urcorner$}}
\newcommand{\code}{\mathsf{code}}
\newcommand{\Det}{\mathsf{Det}}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\oes}{\sim}

@@

We have therefore built an \emph{unbounded nondeterminism} machine, that can print out arbitrarily large natural numbers but which never diverges.  
This is strictly more powerful than finitary choice nondeterminism.
\footnote{Using recursion, we can build a program out of finite nondeterminism that can produce arbitrarily large natural numbers; however, this program also admits the possibility of divergence.}  
What we have just shown is that if we want to solve the problem of building a fair task scheduler, then we must in particular be able to solve the problem of building an unbounded nondeterminism machine.

This is an important observation to make about concurrent programming, because the task of implementing unbounded nondeterminism is difficult -- indeed, considerably more so than that of implementing bounded nondeterminism.  
Dijkstra argues in \cite[Ch. 9]{DijkstraBook} that it is impossible to implement unbounded nondeterminism, showing that the natural constructs from which we construct imperative programs satisfy a \emph{continuity} property that unbounded nondeterminism lacks.  
This is a problem in itself for semanticists, for whom continuity is often a key ingredient in proofs of computational adequacy and full abstraction.

@@

Let $T$ be an Idealized Algol type, and let $M,N\from T$ be closed terms.
Then we write $M\lmam N$ if for all compatible contexts $C[-]$ of ground type we have
\begin{gather*}
  C[M]\converges V \Rightarrow C[N] \converges V\\
  C[M]\mustconverge \Rightarrow C[N] \mustconverge
\end{gather*}

@@

An \emph{arena} is given by a triple $A=(M_A,\lambda_A,\ts_A)$, where
\begin{itemize}
  \item $M_A$ is a countable set of moves,
  \item $\lambda_A\from M_A\to\OP\times\QA$ designates each move as either an \emph{$O$-move} or a \emph{$P$-move}, and as either a \emph{question} or an \emph{answer}.  
    We define $\lambda_A^{OP}=\pr_1\circ\lambda_A$ and $\lambda_A^{QA}=\pr_2\circ\lambda_A$.  
    We also define $\neg\from\OP\times\QA\to\OP\times\QA$ to be the function that reverses the values of $O$ and $P$ while leaving $\QA$ unchanged.
  \item $\ts_A$ is an \emph{enabling relation} between $M_A+\{*\}$ and $M_A$ satisfying the following rules:
    \begin{itemize}
      \item If $a\ts_A b$ and $a\ne b$, then $\lambda_A^{OP}(a)\neq\lambda_A^{OP}(b)$.  
      \item If $*\ts_A a$, then $\lambda_A(a)=OQ$ and $b\nts_A a$ for all $b\in M_A$.
      \item If $a\ts_A b$ and $b$ is an answer, then $a$ is a question.
    \end{itemize}
    We say that a move $a\in M_A$ is \emph{initial} in $A$ if $*\ts_A a$.
\end{itemize}

@@

@@

Harmer and McCusker's approach works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;n;(g\;(\pred m)\;n))\wn\,.
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy $\nu$ with maximal plays of arbitrary length of the form shown in Figure \ref{fig:mn-plays}(b).
Note that this strategy strictly contains the strategy $\mu$ that we considered before, and therefore that the denotation of
  $\IfO \wn M N$
has the same denotation as $N$, even though for any $n$, $M\n\not\mustconverge$, while $N\n\mustconverge$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergent play into the composite strategy -- even though $N$ must converge.

@@

What this illustrates is that we can no longer deduce the infinitary behaviour of a strategy by looking at the limits of its finite plays; instead, we need to keep track of infinite sequences of moves explicitly, in the style of \cite{RoscoeCspInfinite} and \cite{LevyGsInfinite}.
When we use this technique, the denotation of $M$ will contain an infinite sequence, while the denotation of $N$ will contain arbitrarily long finite sequences, but no infinite sequences.

@@

We define an \emph{infinite justified string} in an arena $A$ in the obvious way.  
We say such a string is \emph{recursive} if it corresponds, via the enumeration on $M_A$, to a recursive function $\bN\to\bN$.

@@

Let $A$ be a game.  
A \emph{strategy} $\sigma$ for $A$ is a pair $(T_\sigma,D_\sigma)$, where:
\begin{itemize}
  \item $T_\sigma$ is a non-empty prefix-closed subset of $\overline{P_A}$ such that if $s\in T_\sigma$ is a $P$-position and $sa\in P_A$ then $sa\in T_\sigma$.
  \item $D_\sigma\subset \overline{P_A}$ is a postfix-closed set of plays in $\overline{P_A}$ that either end with an $O$-move or are infinite.  
    We require $D_\sigma$ to obey the following rules:
    \begin{description}
      \item[Divergences come from plays] If $d\in D_\sigma$ then there exists $s\prefix d$ such that $s\in T_\sigma\cap D_\sigma$.
      \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position, then either $s\in D_\sigma$ or $sa\in T_\sigma$ for some $sa$.
      \item[Infinite positions are divergent] If $s\in T_\sigma$ is infinite, then $s\in D_\sigma$.
    \end{description}
\end{itemize}

@@

Note that there is no need to consider separately, as Harmer and McCusker do, divergences that arise through `infinite chattering': in our model, a case of infinite chattering between strategies $\sigma$ and $\tau$ is itself a legal interaction between the two strategies, which is necessarily divergent (because it is infinite) and therefore gives rise to some divergence in $\sigma;\tau$.

@@

\begin{definition}
  Let $\sigma$ be a strategy for a game $A$.  
  We say that $\sigma$ is \emph{locally complete} if it may be written as the union of countably many complete strategies; i.e., there exist $\sigma_n$ such that $T_\sigma=\bigcup T_{\sigma_n}$ and $D_\sigma=\bigcup D_{\sigma_n}$.
\end{definition}

Henceforth, we use `strategy' to mean \emph{locally complete strategy}.

We need to show that the composition of locally complete strategies is locally complete.  
Note that the composition of \emph{complete} strategies is not necessarily complete: for example, our term $N$ above can be written as $N'\;\wn$, where $N'$ is a deterministic term with complete denotation $\nu'$.  
Then we have $\nu=\top_{\bN};\nu'$, but $\nu$ is not complete.
However, we can show that the composition of \emph{deterministic} complete strategies is complete; since a locally complete strategy may always be written as the union of complete deterministic strategies, this is sufficient to show that the composition of locally complete strategies is locally complete.

\begin{definition}
  We say that a strategy $\sigma$ for a game $A$ is \emph{deterministic} if
  \begin{itemize}
    \item it is complete;
    \item if $sab,sac$ are $P$-plays in $T_\sigma$ then $b=c$ and the justifier of $b$ is the justifier of $c$;
    \item if $s\in D_\sigma$ then either $s$ is infinite or there is no $a$ such that $sa\in T_\sigma$.
  \end{itemize}
\end{definition}

@@

@@

Our proof for Corollary \ref{cor:diverge-or-reply} really makes use of the fact that a locally complete strategy is \emph{lively} in the sense of \cite{LevyGsInfinite}; i.e., locally deterministic.  
Our definition is slightly stronger than liveliness, because it insists that the union of complete strategies be countable.
This will be essential to our definability result.

@@

The resulting category $\G^\oc_{D,vis}$ of games and deterministic visible strategies is a fully abstract model of Idealized Algol \cite{SamsonGuyIAPassive}.

@@

If $\sigma$ is locally complete, we say that $\sigma$ is \emph{recursive} if $\sigma$ is the union of complete recursive strategies $\sigma_1,\sigma_2,\cdots$, where the map $i\mapsto \sigma_i$ is a recursive function $\bN \to (\bN \to \bN) \to 2$.

@@

  We first assume that the strategy $\sigma$ is complete.
  Then the strategy $\Det(\sigma)$ is very easy to describe.  
  For each $O$-position $sa\in T_\sigma$, we have some set $B$ of possible replies to $sa$, which we order as $b_1,b_2,\cdots$, where $\code_A(b_1)<\code_A(b_2)<\cdots$.  
  We insert a request to the oracle for a natural number; then, depending on her answer $j$, we play the next move as follows:
  \begin{itemize}
    \item If $0 < j \le \code_A(b_1)$, then play $b_1$.
    \item If $\code_A(b_n) < j \le \code_A(b_{n+1})$ then play $b_{n+1}$.
    \item If $j = 0$ and $sa\in D_\sigma$, then play nothing, and put the resulting play inside $D_{\Det(\sigma)}$.  
      Otherwise, play $b_1$.
  \end{itemize}
  We close under limits to make the strategy $\Det(\sigma)$ complete.  
  $\Det(\sigma)$ is clearly deterministic.  
  Checking that $\top_{\bN};\Det(\sigma)=\sigma$ is easy for finite plays; for infinite plays, it follows by completeness of $\sigma$.

@@

Any term $M\from T$ of Idealized Algol with countable nondeterminism may be written as $M = C[\wn]$, where $C$ is a multi-holed context not involving the constant $\wn$.  
Then the term $\lambda n.C[n]$ is a term of Idealized Algol, and therefore has a denotation $\oc\bN \to \deno{T}$ as in \cite{SamsonGuyIAPassive}.
We define the denotation of $M$ to be given by the composite
\[
  I \xrightarrow{\top_{\bN}}
  \oc \bN \xrightarrow{\deno{\lambda n.C[n]}}
  \deno{T}\,.
  \]
In other words, we interpret the constant $\wn$ using the strategy $\top_{\bN}$ for $\bN$.

@@

@@

For proving full abstraction , it is necessary to take the intrinsic quotient in order to identify, for example, the denotations of $\lambda n.\Omega$ and $\lambda n.\IfO n\;\Omega\;\Omega$ of type $\nat\to\nat$.  
These terms are clearly observationally equivalent, but their denotations are not equal; for example, $q\in D_{\deno{\lambda n.\Omega}}$, but $q\not\in D_{\deno{\lambda n.\IfO n\;\Omega\;\Omega}}$.

@@

Given a nondeterministic strategy $\sigma$ for a game $A$, we may treat $\sigma$ as a game in its own right (a sub-game of $A$).  
Moreover, for any $s\in T_\sigma$, we have a particular branch of that game in which play starts at $s$.  
We say that $s$ is \emph{unreliable} if player $P$ has a strategy for the game starting at $s$ that ensures that play eventually ends up in $D_\sigma$.

@@

The principal difficulties in modelling nondeterministic stateless languages were overcome by Tsukada and Ong in \cite{TsukadaSheaves}, where they outlined how to define an innocent nondeterministic strategy by retaining `branching time information' in strategies.  
An additional benefit of the retention of branching time information is that we no longer need to keep track of infinite plays in order to model unbounded nondeterminism.  
The model given in \cite{TsukadaSheaves} is not sound for must-equivalence, but the authors make the claim that it their model may be easily modified to yield a model that \emph{is} sound for this type of equivalence, using the same techniques from \cite{mcCHFiniteND} that we have used.

>>>>

\acknowledgements{This material is based on work supported by the EPSRC under Grant No.~EP/K018868/1.  
I (Gowers) am grateful to Martin Hyland for our conversation that helped me to develop some of this material.
We are also grateful for the comments made by the anonymous referees for helping to clarify certain points, elaborate on technical difficulties and point to similarities with other parts of the literature.\vspace{24pt}}%optional

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{Dan Ghica and Achim Jung}
\EventNoEds{2}
\EventLongTitle{27th EACSL Annual Conference on Computer Science Logic (CSL 2018)}
\EventShortTitle{CSL 2018}
\EventAcronym{CSL}
\EventYear{2018}
\EventDate{September 4 -- 7, 2018}
\EventLocation{Birmingham, GB}
\EventLogo{}
\SeriesVolume{119}
\ArticleNo{23}
\nolinenumbers %uncomment to disable line numbering
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\nat}{{\mathtt{nat}}}
\newcommand*\from{\colon}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\pred}{{\mathtt{pred\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\com}{{\mathtt{com}}}
\newcommand{\skipp}{{\mathsf{skip}}}
\newcommand{\neww}{{\mathsf{new}}}
\newcommand{\mkvar}{{\mathsf{mkvar}\;}}
\newcommand{\deref}{\texttt{@}}
\newcommand{\ia}[2]{\langle #1 , #2 \rangle}
\newcommand{\stup}[3]{\langle #1 \mid #2 \mapsto #3 \rangle}
\newcommand{\Var}{{\mathtt{Var}}}
\newcommand{\new}{{\mathtt{new}}}
\newcommand{\case}{{\mathtt{case}}}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\bN}{\mathbb{N}}
\newcommand{\bC}{\mathbb{C}}
\newcommand{\deno}[1]{\left\llbracket#1\right\rrbracket}
\newcommand{\converges}{\Downarrow}
\newcommand{\mustconverge}{\converges^{\text{must}}}
\newcommand{\Iflt}{\mathtt{If{<}\;}}
\newcommand{\Ifgt}{\mathtt{If{>}\;}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{\lmam}{\sqsubseteq_{m\&m}}
\newcommand{\emam}{\equiv_{m\&m}}
\newcommand{\lst}{\lesssim}
\newcommand{\rulename}[1]{\LeftTirNameStyle{#1}}
\newcommand{\ts}{{\;\vdash}}
\newcommand{\nts}{{\;\not\vdash}}
\newcommand{\OP}{\{O,P\}}
\newcommand{\QA}{\{Q,A\}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\esuchthat}{\;.\;}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\tensor}{\otimes}
\renewcommand{\implies}{\multimap}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\s}{\mathfrak s}
\newcommand{\C}{{\mathcal{C}}}
\newcommand{\G}{\mathcal G}
\DeclareMathOperator{\Int}{int}
\newcommand{\dv}{{\lightning}}
\DeclareMathOperator{\pocl}{pocl}
\newbox\gnBoxA
\newdimen\gnCornerHgt
\setbox\gnBoxA=\hbox{$\ulcorner$}
\global\gnCornerHgt=\ht\gnBoxA
\newdimen\gnArgHgt
\def\pv #1{%
    \setbox\gnBoxA=\hbox{$#1$}%
    \gnArgHgt=\ht\gnBoxA%
    \ifnum     \gnArgHgt<\gnCornerHgt \gnArgHgt=0pt%
    \else \advance \gnArgHgt by -\gnCornerHgt%
    \fi \raise\gnArgHgt\hbox{$\ulcorner$} \box\gnBoxA %
    \raise\gnArgHgt\hbox{$\urcorner$}}
\newcommand{\code}{\mathsf{code}}
\newcommand{\Det}{\mathsf{Det}}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\oes}{\sim}
\newcommand{\orr}{\textsf{ or }}
\newcommand{\lunit}{\textsf{lunit}}

@@

We have therefore built an \emph{unbounded nondeterminism} machine, that can print out arbitrarily large natural numbers but which never diverges.  
This is strictly more powerful than finitary choice nondeterminism\footnote{Using recursion, we can build a program out of finite nondeterminism that can produce arbitrarily large natural numbers; however, this program also admits the possibility of divergence, unless we are able to insist on fairness.}.
What we have just shown is that if we want to solve the problem of building a fair task scheduler, then we must in particular be able to solve the problem of building an unbounded nondeterminism machine.

This is an important observation to make about concurrent programming, because the task of modelling unbounded nondeterminism is difficult -- indeed, considerably more so than that of modelling bounded nondeterminism.  
Dijkstra argues in \cite[Ch. 9]{DijkstraBook} that it is impossible to implement unbounded nondeterminism, showing that the natural constructs from which we construct imperative programs satisfy a \emph{continuity} property that unbounded nondeterminism lacks.  
Park \cite{park1980semantics} shows that these problems can be surmounted if we use a weaker version of continuity (e.g., $\omega_1$- rather than $\omega$-continuity), but the failure of composition to be continuous is a problem in itself for semanticists, for whom continuity is often a key ingredient in proofs of computational adequacy and full abstraction.

@@

Let $T$ be an Idealized Algol type, and let $M,N\from T$ be closed terms.
Then we write $M\lmam N$ if for all contexts $C[-]$ of ground type with a hole of type $T$, we have
\begin{gather*}
  C[M]\converges V \Rightarrow C[N] \converges V\\
  C[M]\mustconverge \Rightarrow C[N] \mustconverge
\end{gather*}

@@

An \emph{arena} is given by a triple $A=(M_A,\lambda_A,\ts_A)$, where
\begin{itemize}
  \item $M_A$ is a countable set of moves,
  \item $\lambda_A\from M_A\to\OP\times\QA$ designates each move as either an \emph{$O$-move} or a \emph{$P$-move}, and as either a \emph{question} or an \emph{answer}.  
    We define $\lambda_A^{OP}=\pr_1\circ\lambda_A$ and $\lambda_A^{QA}=\pr_2\circ\lambda_A$.  
    We also define $\neg\from\OP\times\QA\to\OP\times\QA$ to be the function that reverses the values of $O$ and $P$ while leaving $\QA$ unchanged.
  \item $\ts_A$ is an \emph{enabling relation} between $M_A+\{*\}$ and $M_A$ satisfying the following rules:
    \begin{itemize}
      \item If $a\ts_A b$, then $\lambda_A^{OP}(a)\neq\lambda_A^{OP}(b)$.  
      \item If $*\ts_A a$, then $\lambda_A(a)=OQ$ and $b\nts_A a$ for all $b\in M_A$.
      \item If $a\ts_A b$ and $b$ is an answer, then $a$ is a question.
    \end{itemize}
    We say that a move $a\in M_A$ is \emph{initial} in $A$ if $*\ts_A a$.
\end{itemize}

@@

If $s$ is a justified string, we will write $sa$ for an arbitrary justified string extending $s$ by a single move $a$, itself justified by some move in $s$.

@@

Harmer and McCusker's approach works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;(n;(g\;(\pred m)\;n)))\wn\,.
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy $\nu$ with maximal plays of arbitrary length of the form shown in Figure \ref{fig:mn-plays}(b).
Note that this strategy strictly contains the strategy $\mu$ that we considered before, and therefore that the denotation of
  $\IfO \wn M N$
has the same denotation as $N$, even though for any $n$, $M\n\not\mustconverge$, while $N\n\mustconverge$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergent play into the composite strategy -- even though $N$ must converge.

@@

What this illustrates is the point made by Park in \cite{park1980semantics} and \cite{park1981concurrency}: namely, that we can no longer deduce the infinitary behaviour of a strategy by looking at the limits of its finite plays; instead, we need to keep track of infinite sequences of moves explicitly, in the style of \cite{RoscoeCspInfinite} and \cite{LevyGsInfinite}.
When we use this technique, the denotation of $M$ will contain an infinite sequence, while the denotation of $N$ will contain arbitrarily long finite sequences, but no infinite sequences.

@@

We define an \emph{infinite justified string} in an arena $A$ in the obvious way.  
We say such a string is \emph{recursive} if it corresponds, via the enumeration on $M_A$, to a pair of recursive functions $\bN\to\bN$ -- one giving the sequence of moves and the other giving the justification relation.

@@

We shall represent a strategy using two sets: a set $T_\sigma$ of \emph{traces}, which takes the role of the plays that may occur in the strategy (as in the usual definition of a deterministic strategy), and a second set $D_\sigma$ of \emph{divergences}; i.e., $O$-positions at which the strategy may elect to diverge.  
In order to model observational equivalence more closely, we shall require $D_\sigma$ to be postfix-closed, since observable contexts cannot detect divergences that occur after the program might already have diverged: consider, for example, the terms $\Omega_{\nat\to\nat}\orr(\lambda n.n)$ and $\Omega_{\nat\to\nat}\orr(\lambda n.(n\orr\Omega_\nat))$ (where we have defined $M\orr N$ to be $\IfO\;\wn\;M\;N$).

For technical reasons we keep track of infinite plays in both $T_\sigma$ and $D_\sigma$, with the rule that any infinite play in $T_\sigma$ must be contained in $D_\sigma$ (since it clearly corresponds to a divergent evaluation).  
We will require that every divergence arise from a trace; i.e., if every play in $D_\sigma$ must have some prefix that is contained in both $T_\sigma$ and $D_\sigma$.  
A consequence of this is that if $d\in D_\sigma$ is infinite and has no finite prefixes in $D_\sigma$, then it must also be contained in $T_\sigma$.  
Not too much importance should be given, however, to the presence or absence of infinite plays in $T_\sigma$: it is quick to show that once we pass to the intrinsic quotient, any such distinction vanishes.

Let $A$ be a game.  
A \emph{strategy} $\sigma$ for $A$ is a pair $(T_\sigma,D_\sigma)$, where:
\begin{itemize}
  \item $T_\sigma$ is a non-empty prefix-closed subset of $\overline{P_A}$ such that if $s\in T_\sigma$ is a $P$-position and $sa\in P_A$ then $sa\in T_\sigma$.
  \item $D_\sigma\subset \overline{P_A}$ is a postfix-closed set of plays in $\overline{P_A}$ that either end with an $O$-move or are infinite.  
    We require $D_\sigma$ to obey the following rules:
    \begin{description}
      \item[Divergences come from traces] If $d\in D_\sigma$ then there exists $s\prefix d$ such that $s\in T_\sigma\cap D_\sigma$.
      \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position, then either $s\in D_\sigma$ or $sa\in T_\sigma$ for some $sa$.
      \item[Infinite positions are divergent] If $s\in T_\sigma$ is infinite, then $s\in D_\sigma$.
    \end{description}
\end{itemize}

@@

Note that there is no need to consider separately, as Harmer and McCusker do, divergences that arise through `infinite chattering': in our model, we will see that a case of infinite chattering between strategies $\sigma$ and $\tau$ is itself a legal interaction between the two strategies, which is necessarily divergent (because it is infinite) and therefore gives rise to some divergence in $\sigma;\tau$.

@@

\begin{definition}
  Let $\sigma$ be a strategy for a game $A$.  
  We say that $\sigma$ is \emph{locally complete} if it may be written as the union of complete strategies; i.e., there exist $\sigma_i$ such that $T_\sigma=\bigcup T_{\sigma_i}$ and $D_\sigma=\bigcup D_{\sigma_i}$.  
  Note that since $T_\sigma$ and $D_\sigma$ are countable sets (because there are countably many recursive plays), this union may be taken to be countable.
\end{definition}

It will be slightly more convenient to use an equivalent definition, based on unions of \emph{deterministic} strategies, which are a special case of complete strategies.

\begin{definition}
  We say that a strategy $\sigma$ for a game $A$ is \emph{deterministic} if
  \begin{itemize}
    \item it is complete;
    \item if $sa,sb$ are $P$-plays in $T_\sigma$ then $a=b$ and the justifier of $a$ is the justifier of $b$;
    \item if $s\in D_\sigma$ then either $s$ is infinite or there is no $a$ such that $sa\in T_\sigma$.
  \end{itemize}
\end{definition}

We say that a strategy $\sigma$ is \emph{lively} or \emph{locally deterministic} if there exists a collection of deterministic strategies $\sigma_i$ such that $T_\sigma=\bigcup T_{\sigma_i}$ and $D_\sigma=\bigcup D_{\sigma_i}$.  
It is clear that a strategy is lively if and only if it is locally complete, and that the collection of $\sigma_i$ may again be taken to be countable.

From now on, we will use \emph{strategy} to mean \emph{lively} (or \emph{locally complete}) \emph{strategy}. This means that we will need to show that the composition of lively strategies is again lively.

@@

It is, of course, true that the composition of deterministic strategies is deterministic, but we do not really need this fact.

@@

@@

After passing to the intrinsic quotient, the resulting category $\G^\oc_{D,vis}$ of games and deterministic visible strategies is a fully abstract model of Idealized Algol \cite{SamsonGuyIAPassive}.

@@

If $\sigma$ is lively, we say that $\sigma$ is \emph{recursive}, and if $\sigma$ is the union of complete recursive strategies $\sigma_1,\sigma_2,\cdots$, where the map $i\mapsto \sigma_i$ is a recursive function $\bN \to (\bN \to \bN) \to 2$.

Note that there are plenty of strategies that we want that are not the union of a recursive sequence of \emph{deterministic} strategies -- for example, the strategy on $(\bN\to\bC)\to\bC$ that calls its natural-number argument infinitely many times is complete and has no $O$-branching, but its infinite traces include every recursive sequence of natural numbers.

Using these definitions, it seems to be hard to prove that the composition of recursive strategies is itself a recursive strategy: the tricky point is to show that the decomposition into complete strategies may still be taken to be given by a recursive strategy.  
The example in the previous paragraph shows that we cannot use the same proof as we did in the non-recursive case, which used deterministic strategies.  
Fortunately, we do not need to be able to show that the composition of recursive strategies is recursive in order to prove our full abstraction result, so we leave this problem for future work.

@@

  We first assume that the strategy $\sigma$ is complete.
  Then the strategy $\Det(\sigma)$ is very easy to describe.  
  For each $O$-position $s\in T_\sigma$, we have some set $B$ of possible replies to $s$, which we order as $b_1,b_2,\cdots$, where $\code_A(b_1)<\code_A(b_2)<\cdots$.  
  We insert a request to the oracle for a natural number; then, depending on her answer $j$, we play the next move as follows:
  \begin{itemize}
    \item If $0 < j \le \code_A(b_1)$, then play $b_1$.
    \item If $\code_A(b_n) < j \le \code_A(b_{n+1})$ then play $b_{n+1}$.
    \item If $j = 0$ and $s\in D_\sigma$, then play nothing, and put the resulting play inside $D_{\Det(\sigma)}$.  
      Otherwise, play $b_1$.
  \end{itemize}
  We close under limits to make the strategy $\Det(\sigma)$ complete.  
  $\Det(\sigma)$ is clearly deterministic.  
  Checking that $\top_{\bN};\Det(\sigma)=\sigma$ is easy for finite plays; for infinite plays, it follows by completeness of $\sigma$.

@@

We model the language compositionally, using denotations as in \cite{mcCHFiniteND} for the nondeterministic constants and modelling $\wn$ using the strategy $\top_{\bN}\from \bN$.

Any term $M\from T$ of Idealized Algol with countable nondeterminism may be written as $M = C[\wn]$, where $C$ is a multi-holed context not involving the constant $\wn$.  
Then the term $\lambda n.C[n]$ is a term of Idealized Algol, and therefore has a denotation $\oc\bN \to \deno{T}$ as in \cite{SamsonGuyIAPassive}.

\begin{lemma}
  The term $C[\wn]$ has the same denotation as the term $(\lambda n.C[n])\wn$.
\end{lemma}

\begin{proof}
  This is a straightforward argument by structural induction on $C$, and the constant $\top_\bN$ does not really play a role.  
  We prove inductively on $T$ that if $\Gamma\ts C[\wn]\from T$ is a term-in-context, then its denotation may be given by the following composite.
  \[
    \deno{\Gamma}
    \xrightarrow{\lunit;(\top_{\bN}\times\id)}
    \bN\times\deno{\Gamma}
    \xrightarrow{\deno{\Gamma,n\from\nat\ts C[n]\from T}}
    \deno{T}
    \]
\end{proof}

@@

Corollary \ref{cor:adequacy} is very general, and is intended to be applied in multiple situations.  
In particular, it may be applied to a game semantics in which we define a `nondeterministic visible strategy' on a game $A$ to be a deterministic visible strategy for $\bN\to A$, up to a suitable equivalence relation.  
This model is an example of a much more general construction that is the subject of ongoing research by the authors.  
In this sense, our main model based on nondeterministic strategies is not necessary in order to obtain our full abstraction result.  
Nevertheless, we felt it important to give a model based on nondeterministic strategies, since these are the `natural' game semantic interpretation of nondeterminism.

@@

For full abstraction, we need to take the intrinsic quotient in order to identify, for example, the terms $\lambda n.\Omega$ and $\lambda n.\IfO n\;\Omega\;\Omega \from \nat\to\nat$.  
These terms are observationally equivalent, but their denotations are not equal; for example, $q\in D_{\deno{\lambda n.\Omega}}$, but $q\not\in D_{\deno{\lambda n.\IfO n\;\Omega\;\Omega}}$.

@@

Given a nondeterministic strategy $\sigma$ for a game $A$, we may treat $\sigma$ as a game in its own right (a sub-game of $A$).  
Moreover, for any $s\in T_\sigma$, we have a particular branch of that game in which play starts at $s$.  
We say that $s$ is \emph{unreliable} if player $P$ has a strategy for the game starting at $s$ that ensures that the (possibly infinite) limiting play is in $D_\sigma$.

@@

The principal difficulties in modelling nondeterministic stateless languages were overcome by Tsukada and Ong in \cite{TsukadaSheaves}, where they outlined how to define an innocent nondeterministic strategy by retaining `branching time information' in strategies.  
An additional benefit of the retention of branching time information is that we no longer need to keep track of infinite plays in order to model unbounded nondeterminism.  
Tsukada and Ong's primary model was based on sheaves over a site of plays, but they also give a more direct way of characterizing nondeterministic innocence, based on ideas by Levy \cite{levy2014morphisms}.
